The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-02-04 Paper.Rmd
various changes

A third reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect copula structure in environmental variables 
to take the form of strong tail dependence if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case.
Spatial dependence in an environmental variable 
tends to beget dependence across space
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This process is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of spatial synchrony.
Synchrony is a phenomenon of great interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

@@

Having demonstrated that non-N copula structure and asymmetric tail dependence are common
features of ecological data, we explore via mathematical 
models two possible mechanisms of this structure which 
seem likely to operate in population systems. The first mechanism 
is an extension of the well known Moran
effect, and was summarized in the Introduction. 
It explores the idea that non-N copula structure and asymmetric tail dependence 
between environmental variables measured in two places may bring about similarly 
structured dependence between ecological variables that are influenced by the 
environmental variables. Let $P_i(t)$ be a population index for location $i$ ($i=1,2$),
and assume the index follows AR(1) dynamics within a patch, 
$P_i(t+1)=\beta P_{i}(t-1)+ \sqrt{1-\beta^2} \epsilon_i(t)$ where the $\epsilon_i(t)$ are 
standard-normal random variables that are independent for distinct times, $t$, but correlated

>>>>

A third reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect copula structure in environmental variables 
to take the form of strong tail dependence if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case.
Spatial dependence in an environmental variable 
tends to beget dependence across space
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This process is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of spatial synchrony of population, community, 
biogeochemical and other ecological variables.
Synchrony is a phenomenon of great interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

@@

Having demonstrated that non-N copula structure and asymmetric tail dependence are common
features of ecological data, we explore via mathematical 
models two possible mechanisms of this structure which 
seem likely to operate in diverse ecological scenarios. The first mechanism 
is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
It explores the idea that non-N copula structure and asymmetric tail dependence 
between environmental variables measured in two places may bring about similarly 
structured dependence between ecological variables that are influenced by the 
environmental variables. Let $P_i(t)$ be a population index for location $i$ ($i=1,2$),
and assume the index follows AR(1) dynamics within a patch, 
$P_i(t+1)=\beta P_{i}(t-1)+ \sqrt{1-\beta^2} \epsilon_i(t)$. Here the $\epsilon_i(t)$ are 
standard-normal random variables that are independent for distinct times, $t$, but 
that are non-independent across patches.

<!--***DAN: Note to self, this model, and also the other one we study in this section,
are actually much more general than populatin. Should replace P_i with E_i where E_i is
just some ecological variable. Could be deviaitons from a pop from carrying capacity, could be a 
biogeochemical variable, could maybe even be an evolutinary variable. Anything that is
affected by an environmental variable and could reasonably be modelled by an AR process. 
So the generalization to biogeochemical and community
data seems pretty straightforward. Need to think this through and reword the corresponding
part of the intro to encompass this greater generality. How would the evolutionary case work?
Well, suppose the two "patches" are actually traits, and the perturbations epsilon are somehow
the "drift events" or something like that? This does not really make sense, strictly speaking,
now that I am thinking about it, but maybe the model could be modified to make it make sense?
At least I have straightforward generalization to biogeochem and community cases. 
More on evolutionary: suppose we model traits as a bivariate random walk where the perturbations
in the traits are related with a copula with some tail dependence. Suppose you have a tree and 
you model evolution down the tree in this way and then look at the bivariate distribution of
trait on the tips of the tree. What do you get? I guess you get some copula structure?-->
